Analysis

### 2022-07-19 manuscript/manuscript_bioinformatics.Rmd
Add Eric's initial
Add Mike's comments

Mikhail G. Dozmorov^1,2,\*^, Wancen Mu^3^, Eric Davis^5^, Stuart Lee^13,14^, Tim Triche^10,11,12^, Douglas Phanstiel^5,6,7,8,9^, Michael Love^3,4^

@@

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transcription regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa; @Ohlsson:2001wu; @Kim:2007ux]. <!--Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv],--> Tt has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, cancer and developmental disorders, chromatin interactions in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy]. <!--CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs]. Mutations in CTCF binding sites and CTCF itself are associated with human cancers [@Liu:2019vg; @Saldana-Meyer:2019aa], identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].-->
<!-- Most CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux]. binding motif length varying between 20 and 40bp -->
CTCF binds DNA through the combinatorial use of its 11-zinc finger domains to target sites with remarkable sequence variation [@Ohlsson:2001wu; @Jolma:2013vd]<!--@Filippova:1996tv-->. A 15-20bp CTCF consensus motif from the ChIP-seq data analysis has been defined [@Kim:2007ux], referred to hereafter as M1. This motif interacts with the N-terminal zinc fingers for most CTCF–DNA-binding events. Subsequently, a shorter 9bp motif (M2) interacting with the C-terminal fingers was discovered that, together with M1, forms CTCF binding sites spanning 34-35bp [@Schmidt:2012uk]. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, and LM23 motifs as parts of CTCF binding.<!--Subsequently, $\sim80\%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences, and ZFs 9-11 were associated with a second conserved motif [@Nakahashi:2013uw; @Pugacheva:2020wz].--> Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes<!--[@Kim:2007ux; @Nakahashi:2013uw; @Wang:2012ty]-->.
<!-- CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa; @Oti:2016aa; @Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa; @Fudenberg:2016aa; @Goloborodko:2016vz; @Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv; @Rao:2014aa; @Vietri-Rudan:2015tu; @Beagan:2020wz; @Stilianoudakis:2020aa; @Pugacheva:2020wz]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops -->

@@

The FIMO tool is one of the oldest and most well-known tool for motif scanning. It uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values [@Cuellar-Partida:2012wj], with the default p-value cutoff 1E-4. However, we observed that up to 98% of CTCF sites are detected within the 1E-4 - 1E-6 p-value range (`Figure 1B`, `Figure S2B`), suggesting most CTCF sites have a relatively poor sequence match. Using the ENCODE SCREEN database of CTCF-bound cis-regulatory elements (cCREs) [@ENCODE-Project-Consortium:2020uq] as ground truth (`Table S2`, included in the package), we quantified the proportion of true and false positives depending on the FIMO p-value cutoff. We found the 1E-6 threshold the most optimal, providing $\sim80\%$ of true positive CTCF motifs (`Figure 1C`, `Figure S2C`). Additionally, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S3`). Reducing them (merging overlapping CTCF sites), combined with 1E-6 cutoff filtering, yields the number of CTCF sites comparable to previously reported ones. Although we provide FIMO results using the default settings, we recommend filtering and reducing them.

>>>>

Mikhail G. Dozmorov^1,2,\*^, Wancen Mu^3^, Eric S. Davis^5^, Stuart Lee^13,14^, Tim Triche^10,11,12^, Douglas Phanstiel^5,6,7,8,9^, Michael I. Love^3,4^

@@

The structural and regulatory organization of the mammalian genome is fundamentally dependent on CTCF (CCCTC-binding factor), a versatile transcription regulator evolutionary conserved from fruit fly to human [@Phillips:2009aa; @Ohlsson:2001wu; @Kim:2007ux]. <!--Originally defined as a repressor/insulator protein [@Lobanenkov:1990vm; @Filippova:1996tv],--> It has been found to be involved in a variety of regulatory functions including transcriptional activation, imprinting, X-chromosome activation, cancer and developmental disorders, chromatin interactions in three dimensions [@Phillips:2009aa; @Ohlsson:2010uy]. <!--CTCF is ubiquitously expressed, and its deletion in the germline cells is incompatible with cell viability [@Heath:2008vs]. Mutations in CTCF binding sites and CTCF itself are associated with human cancers [@Liu:2019vg; @Saldana-Meyer:2019aa], identifying CTCF as an important candidate tumor-suppressor gene [@Filippova:2002uz].-->
<!-- Most CTCF binding sites are located far from the transcriptional start sites, with their distribution strongly correlated with genes [@Kim:2007ux]. binding motif length varying between 20 and 40bp -->
CTCF binds DNA through the combinatorial use of its 11-zinc finger domains to target sites with remarkable sequence variation [@Ohlsson:2001wu; @Jolma:2013vd]<!--@Filippova:1996tv-->. A 15-20bp CTCF consensus motif from the ChIP-seq data analysis has been defined [@Kim:2007ux], referred to hereafter as M1. This motif interacts with the N-terminal zinc fingers for most CTCF–DNA-binding events. Subsequently, a shorter 9bp motif (M2) interacting with the C-terminal fingers was discovered that, together with M1, forms CTCF binding sites spanning 34-35bp [@Schmidt:2012uk]. The complexity of CTCF binding was also noted in the earlier study of motif discovery in conserved noncoding elements by Xie et al. [@Xie:2007wm], detecting LM3, LM7, and LM23 motifs as parts of CTCF binding.<!--Subsequently, $\sim80\%$ of CTCF conserved motifs were found to be bound by ZFs 4-7, while ZFs 1-2 and 8-11 bind to nonconserved flanking sequences, and ZFs 9-11 were associated with a second conserved motif [@Nakahashi:2013uw; @Pugacheva:2020wz].--> Studies have reported 20,000-50,000 CTCF binding sites within the human and mouse genomes<!--[@Kim:2007ux; @Nakahashi:2013uw; @Wang:2012ty]-->.
<!-- CTCF binding has long been known to be critical for Topologically Associating Domain (TAD) and chromatin loop formation [@Dixon:2012aa; @Oti:2016aa; @Vietri-Rudan:2015aa]. Loops are formed via cohesin-mediated loop extrusion mechanism [@Alipour:2012aa; @Fudenberg:2016aa; @Goloborodko:2016vz; @Schwarzer:2016aa] with CTCF (in convergent orientation) and cohesin marking domain boundaries [@Nishana:2020tv; @Rao:2014aa; @Vietri-Rudan:2015tu; @Beagan:2020wz; @Stilianoudakis:2020aa; @Pugacheva:2020wz]. To demonstrate the use of CTCF data package, we provide a Python script that quantifies the proportion of loops -->

@@

The FIMO tool is one of the oldest and most well-known tool for motif scanning. It uses a dynamic programming algorithm to convert log-odds scores of motif matching into p-values [@Cuellar-Partida:2012wj], with the default p-value cutoff 1E-4. However, we observed that up to 98% of CTCF sites are detected within the 1E-4 - 1E-6 p-value range (`Figure 1B`, `Figure S2B`), suggesting most CTCF sites have a relatively poor sequence match. Using the ENCODE SCREEN V3 database of CTCF-bound cis-regulatory elements (cCREs) [@ENCODE-Project-Consortium:2020uq] as ground truth (`Table S2`, included in the package), we quantified the proportion of true and false positives depending on the FIMO p-value cutoff. We found the 1E-6 threshold the most optimal, providing $\sim80\%$ of true positive CTCF motifs (`Figure 1C`, `Figure S2C`). Additionally, given some databases provide multiple CTCF PWMs, FIMO detects many overlapping CTCF sites (`Table S3`). Reducing them (merging overlapping CTCF sites), combined with 1E-6 cutoff filtering, yields the number of CTCF sites comparable to previously reported ones. Although we provide FIMO results using the default settings, we recommend filtering and reducing them.
